---
output: html_document
---
Analysis of MPG vs Transmission type
====================================

Alper Halbutogullari, November 10, 2015

## Executive summary

We are trying to answer the following questions:

*“Is an automatic or manual transmission better for MPG”*   
*"Quantify the MPG difference between automatic and manual transmissions"*  

Based on the given (mtcars) data set, all else being equal,
cars with manual transmissions are more efficient than the ones with
automatic transmissions. The difference is, on average, 7.24 mpg.
This value has a 95% confidence interval of [3.21 , 11.28],
in other words we are 95% confident that the savings is 7.24 $\pm$ 4.03 mpg.

## Analysis

### Summary of the data

The data was extracted from the 1974 Motor Trend US magazine, and
comprises fuel consumption and other aspects of automobile design and
performance for 32 automobiles (1973–74 models). We will only be using
the mpg & am (transmission type, 0 = Automatic, 1 = Manual) features.

```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=6}
data(mtcars)
```

### Initial Exploration

If we visualize how mpg changes with transmission type (see Figure 1 in Appendix), then we see that there is clearly a difference (~7 mpg). However we need to check the significance of this hypothesis.

Although transmission type is a categorical variable, because it has
only two values, we can treat it as numerical. So let's check the
correlation between mpg and the transmission type:

```{r echo=TRUE, eval=TRUE, fig.width=10, fig.height=3}
cor.test(mtcars$mpg, mtcars$am)
```

Based on the p value (0.000285) from the above result, 
we can confidently say that there is a direct correlation.

So, let's try linear regression to model the relation
between MPG and the transmission type:

```{r echo=TRUE, eval=TRUE, fig.width=10, fig.height=3}
fit <- lm(mpg ~ am, data = mtcars) 
summary(fit)
```

This model tells us that we can formulate the mpg-am relation 
by the following formula (see Figure 2 for more details) :

    mpg = 17.147 + 7.245 * am

7.245, the coefficient of am in the above formula, is the mpg
difference between cars with automatic and manual transmissions
(with automatic transmissions consuming more gallons of gas per mile).

Let's also check the significance of this with a t-test:

```{r echo=TRUE, eval=TRUE, fig.width=10, fig.height=3}
t.test(mpg ~ am, data=mtcars)
```

From the result of the above t-test, we see that mpg values of the cars
with manual and automatic transmission are 17.14737 and 24.39231,
respectively, hence the difference is: 7.24494. 
From the p-value (0.001374 << 0.05), we can confidently say that this
difference is significant, i.e. not happening by chance. Also from the
95% confidence interval, we can say that this difference varies
between 3.209684 and 11.280194. Note that the result of this test 
agrees exactly with the result of the regression model.

## Conclusion.

As seen from both of the results above, cars with manual
transmission provide a savings of, on average, 7.24494 mpg 
over cars with automatic transmission, with a 95% confidence 
interval of [3.21 , 11.28].


## Appendix


```{r warning=FALSE, echo=FALSE, eval=TRUE, fig.width=10, fig.height=10}
library(ggplot2)

mtcars$am<-factor(mtcars$am, levels=c(0,1), labels=c("Automatic", "Manual"))

ggplot(mtcars, aes(y=mpg, x=factor(am))) +
       geom_boxplot(aes(fill = factor(am))) +
       xlab("Transmission type") +
       ylab("Miles per gallon") +
       labs(title=expression("MPG vs. transmission type"))
```


Figure 1: How mpg changes with transmission type



```{r warning=FALSE, echo=FALSE, eval=TRUE, fig.height=8}
par(mfrow=c(2,2))
plot(fit)
```


Figure 2: Details of model "fit"
